Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2023-02-06 manuscript.Rmd
remove UTF8 chars

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either it was observed or not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with the help of the Bayes formula:

@@

The posterior probability ($P(H|E)$) is of great relevance since it is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (XXX), in its direct use, it is called corroboration as firmness; in its relation to the a priori probability ($P(H)$), it is called increase in firmness.
As noted before, we concentrate on posterior probability as a measure of epistemic value, since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in posterior probability ($P(H|E)$).
We assume that any measure shows gain if, and only if, evidence is observed assuming $P(E|H) > P(E|¬H)$.
This assumption is sometimes called statistical relevance condition (XXX).

@@

\begin{equation}
P(E) = P(H)P(E|H) + P(¬H)P(E|¬H) (\#eq:total-expectedness)
\end{equation}

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(¬H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
@kuklaClinicalStatisticalTheory1990 argued that the core arguments in @meehlAppraisingAmendingTheories1990 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ can be related to or even equated to theoretical risk.

Let us note some interesting properties of theoretical risk ($P(¬E|¬H)$).
First, increasing theoretical risk leads to higher posterior probability ($P(H|E)$, our objective).
Second, if the theoretical risk is smaller than detectability ($P(E|H)$) it follows that the posterior probability must decrease when observing the evidence.
If detectability exceeds theoretical risk, the evidence is less likely under the theory than it is when the theory does not hold.
Third, if the theoretical risk equals zero, then posterior probability is at best equal to prior probability but only if detectability is perfect ($P(H|E)$ = 1).
In other words, observing a sure fact does not lend credence to a hypothesis.

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(¬E|¬H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

Both theoretical risk ($P(¬E|¬H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories.
To illustrate the concepts we introduced up to here, consider the following  example of a single theory and three experiments that may test it. 
The experiments were created to illustrate how they may differ in their theoretical risk and detectability.
Suppose the primary theory is about the cognitive phenomenon of "insight."
For the purpose of illustration, we define it, with quite some hand-waving, as a cognitive abstraction that allows agents to consistently solve a well-defined class of problems.
We present the hypothesis that the following problem belongs to such a class of insight problems:

@@

We argue that experiment 1 has high theoretical risk ($P(¬E_1|¬H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($¬H$), then presenting the insight that Roman numerals could be used, should not lead to all students solving the problem ($¬E_1$); the experiment therefore has high theoretical risk ($P(¬E_1|¬H)$).
Conversely, if insight is required to solve the problem ($H$), then it is likely to help all students to solve the problem ($E_1$), the experiment therefore has high detectability ($P(E_1|H)$).
The second experiment, on the other hand, has low theoretical risk ($P(¬E_2|¬H)$).
Even if "insight" has nothing to do with solving the problem ($¬H$), there are other plausible reasons for observing the evidence ($E_2$), because the students could simply copy the solution, without having any insight.
With regard to detectability, experiments 1 and 2 differ in no obvious way.
Experiment 3, however, also has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$) experiment 3 therefore has low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms, but high compared to the detectability.
In the unlikely event that all 10 students place their matches to form the Roman numeral Ⅷ ($E_3$), it is probably due to insight ($H$) and not by chance $P(¬E_3|¬H)$).
Of course, in practice, we would allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.

@@

Second, available resources limit theoretical risk.
Increasing theoretical risk ($P(¬E|¬H)$) will usually decrease detectability ($P(E|H)$) unless more resources are invested.
In other words, one cannot increase power while maintaining the same type-I error rate without increasing the invested resources.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.

@@

Theoretical risk is deeply connected with statistical methods, because it is the inverse of $P(E|¬H)$.
$P(E|¬H)$ is equivalent to the type-I error rate, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, because then the nill hypothesis is $¬H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, the research hypothesis seldomly equals  the statistical null hypothesis, and therefore, the relation between statistical type-I error rate and theoretical risk should not be overinterpreted.
We argue that theoretical risk (and hence its inverse, $P(E|¬H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

Statistical methods stand out among these factors because we have a large toolbox for assessing and controlling their contribution to theoretical risk.
Examples of our ability to exert this control are the setting of type-I error rate, the use of corrected fit measures (i.e., adjusted R²), information criteria, or cross-validation in machine learning.
These tools help us account for biases in statistical methods that increase the likelihood of signifying results even when there are none ($P(E|¬H)$).

@@

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test  of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden, enlarge their definition of the population to be sampled, translate their questionnaires into a different language, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|¬H)$.
While, in theory, these factors may leave $P(E|¬H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.

>>>>

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either it was observed or not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(\neg{}H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with the help of the Bayes formula:

@@

The posterior probability ($P(H|E)$) is of great relevance since it is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (XXX), in its direct use, it is called corroboration as firmness; in its relation to the a priori probability ($P(H)$), it is called increase in firmness.
As noted before, we concentrate on posterior probability as a measure of epistemic value, since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in posterior probability ($P(H|E)$).
We assume that any measure shows gain if, and only if, evidence is observed assuming $P(E|H) > P(E|\neg{}H)$.
This assumption is sometimes called statistical relevance condition (XXX).

@@

\begin{equation}
P(E) = P(H)P(E|H) + P(\neg{}H)P(E|\neg{}H) (\#eq:total-expectedness)
\end{equation}

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(\neg{}H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|\neg{}H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(\neg{}E|\neg{}H)= 1 - P(E|\neg{}H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(\neg{}E|\neg{}H)$.
@kuklaClinicalStatisticalTheory1990 argued that the core arguments in @meehlAppraisingAmendingTheories1990 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(\neg{}E|\neg{}H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|\neg{}H)$ which would imply that $P(\neg{}E|\neg{}H)$ can be related to or even equated to theoretical risk.

Let us note some interesting properties of theoretical risk ($P(\neg{}E|\neg{}H)$).
First, increasing theoretical risk leads to higher posterior probability ($P(H|E)$, our objective).
Second, if the theoretical risk is smaller than detectability ($P(E|H)$) it follows that the posterior probability must decrease when observing the evidence.
If detectability exceeds theoretical risk, the evidence is less likely under the theory than it is when the theory does not hold.
Third, if the theoretical risk equals zero, then posterior probability is at best equal to prior probability but only if detectability is perfect ($P(H|E)$ = 1).
In other words, observing a sure fact does not lend credence to a hypothesis.

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

Both theoretical risk ($P(\neg{}E|\neg{}H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories.
To illustrate the concepts we introduced up to here, consider the following  example of a single theory and three experiments that may test it. 
The experiments were created to illustrate how they may differ in their theoretical risk and detectability.
Suppose the primary theory is about the cognitive phenomenon of "insight."
For the purpose of illustration, we define it, with quite some hand-waving, as a cognitive abstraction that allows agents to consistently solve a well-defined class of problems.
We present the hypothesis that the following problem belongs to such a class of insight problems:

@@

We argue that experiment 1 has high theoretical risk ($P(\neg{}E_1|\neg{}H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($\neg{}H$), then presenting the insight that Roman numerals could be used, should not lead to all students solving the problem ($\neg{}E_1$); the experiment therefore has high theoretical risk ($P(\neg{}E_1|\neg{}H)$).
Conversely, if insight is required to solve the problem ($H$), then it is likely to help all students to solve the problem ($E_1$), the experiment therefore has high detectability ($P(E_1|H)$).
The second experiment, on the other hand, has low theoretical risk ($P(\neg{}E_2|\neg{}H)$).
Even if "insight" has nothing to do with solving the problem ($\neg{}H$), there are other plausible reasons for observing the evidence ($E_2$), because the students could simply copy the solution, without having any insight.
With regard to detectability, experiments 1 and 2 differ in no obvious way.
Experiment 3, however, also has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$) experiment 3 therefore has low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms, but high compared to the detectability.
In the unlikely event that all 10 students place their matches to form the Roman numeral VIII ($E_3$), it is probably due to insight ($H$) and not by chance $P(\neg{}E_3|\neg{}H)$).
Of course, in practice, we would allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.

@@

Second, available resources limit theoretical risk.
Increasing theoretical risk ($P(\neg{}E|\neg{}H)$) will usually decrease detectability ($P(E|H)$) unless more resources are invested.
In other words, one cannot increase power while maintaining the same type-I error rate without increasing the invested resources.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.

@@

Theoretical risk is deeply connected with statistical methods, because it is the inverse of $P(E|\neg{}H)$.
$P(E|\neg{}H)$ is equivalent to the type-I error rate, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, because then the nill hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, the research hypothesis seldomly equals  the statistical null hypothesis, and therefore, the relation between statistical type-I error rate and theoretical risk should not be overinterpreted.
We argue that theoretical risk (and hence its inverse, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

Statistical methods stand out among these factors because we have a large toolbox for assessing and controlling their contribution to theoretical risk.
Examples of our ability to exert this control are the setting of type-I error rate, the use of corrected fit measures (i.e., adjusted R²), information criteria, or cross-validation in machine learning.
These tools help us account for biases in statistical methods that increase the likelihood of signifying results even when there are none ($P(E|\neg{}H)$).

@@

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test  of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden, enlarge their definition of the population to be sampled, translate their questionnaires into a different language, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|\neg{}H)$.
While, in theory, these factors may leave $P(E|\neg{}H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.
